---
title: "Exercises 1"
date: "22/11/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Ordinary least squares

We begin by generating $n=100$ equally spaced points in the interval between $0$ and $\pi$. Denote the points $\tau_i$ for $i=1,\dots,100$.
```{r}
# Set the number n
nn <- 100
# Generate the points
tau <- seq(from = 0, to = pi, length.out = nn)
```

We can inspect the elements of `tau` by simply calling the variable.
```{r}
tau
```

We wish to generate the matrix $\bf{X}$, where the $i$th row of $\bf{X}$ is
$$\bar{\boldsymbol{x}}_{i}^{\top}=\left(1,\sin\left(\frac{\pi\tau_{i}}{2}\right),\cos\left(\frac{\pi\tau_{i}}{2}\right),\sin\left(\pi\tau_{i}\right),\cos\left(\pi\tau_{i}\right)\right).$$
```{r}
# Begin by generating a column of 1s
XX <- cbind(rep(1,nn))
# Make a for loop to bind columns with cos and sin values of increasing frequency
for (kk in 1:2) {
  XX <- cbind(XX,sin(pi/2*tau*kk),cos(pi/2*tau*kk))
}
# Inspect the first 5 rows of XX
head(XX)
```

Set $\alpha=1$ and $\boldsymbol{\beta}^\top=(1,1,1,1)$, and $\boldsymbol{\theta}^\top=(\alpha,\boldsymbol{\beta}^\top)$. Then, for each $i$, generate
$$y_{i}=\boldsymbol{\theta}^{\top}\bar{\boldsymbol{x}}_{i}+u_{i},$$
where $u_i$ is normally distributed with mean $0$ and variance $1/4$.


